---
title: "Food Consumption and CO2 Emissions, Tidy Tuesday"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

Sources

Data from Tidy Tuesday (<https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-18/readme.md>), February 18, 2020. I started by watching and following along with Andrew Couch's EDA on YouTube (<https://www.youtube.com/watch?v=VKCPYet9qLM>).

Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r, include=FALSE}
library(tidytuesdayR)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(broom)
library(reactable)
library(readr)
library(kableExtra)
library(plotly)
```

Data

```{r, include=FALSE}
tuesdata <- tidytuesdayR::tt_load('2020-02-18')
tuesdata <- tidytuesdayR::tt_load(2020, week = 8)


food_consumption <- tuesdata$food_consumption %>%
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE))) %>%
rename(co2_emission = co2_emmission)

```

| Variable      | Class     | Description                       |
|---------------|-----------|-----------------------------------|
| country       | character | Country Name                      |
| food_category | character | Food Category                     |
| consumption   | double    | Consumption (kg/person/year)      |
| co2_emission  | double    | Co2 Emission (kg CO2/person/year) |

: Data Dictionary

Variables of interest

```{r}
#Review the data
#130 countries and 11 food categories
food_consumption %>% count(country) 

#basic summary of the data set
food_consumption_NA <- food_consumption |>
  summarize(across(everything(), 
                   \(x) mean(is.na(x))))

#No NAs
food_consumption_0 <- 
  nrow(food_consumption[food_consumption$consumption == "0",]) #31 cases

#which countries/categories have cases where consumption = 0?
countries_0consumption <- food_consumption %>%
  filter(consumption == "0") %>%
  group_by(country) %>%
  print(n=31)
#food categories with a "0" value: Soybeans (22); Pork (8), and Lamb & Goat (1).
#Pork makes a lot of sense - countries with large Muslim populations are included in this list
food_consumption %>% 
  filter(consumption == "0" & food_category == "Pork") %>%
  group_by(country) %>%
  arrange(country) %>% 
  print()
#but what about soybeans? While I can believe that some of these countries eat very few soy products, I have a hard time believing they eat none. I am sceptical of these 0's.
food_consumption %>% 
  filter(consumption == "0" & food_category == "Soybeans") %>%
  group_by(country) %>%
  arrange(country) %>% 
  print(n=Inf)
```

Visualize

1. Boxplot showing log distribution of consumption and co2_emission. From Andrew Couch's video.
```{r}
#Distributions of CO2 and consumption are not as proportional as it appeared at first
options(scipen = 999)
food_consumption %>%
#tidy up the data before creating visual
  gather(key = "feature", value = "value", -country, -food_category) %>%
  ggplot(aes(x=feature, y = value, color = feature)) + 
  geom_boxplot() +
  geom_jitter(alpha = 0.1)+
  facet_wrap(~feature, scales = "fixed")+
  scale_y_log10()
```

2. Bar chart showing average the average energy efficiency of each food category. Meat products like lamb & goat and beef are high CO 2 emitters per kg consumed, wheat and wheat products as well as soybeans are the most energy efficient.
```{r}
#Consumption = kg eaten by each person/year
#Co2 emission = kg of co2 produced by each person/year

#AC video question: for every kg of CO2 emitted, how many kgs of food do they eat? He's trying to determine which countries are more efficient. 
#of kg of food eaten produces how many kgs of CO2?

#Are more developed countries consuming more food / less efficient?

food_consumption %>%
  filter(consumption > 0 & co2_emission >0) %>%
  mutate(co2perfood = co2_emission/consumption) %>%
  group_by(food_category) %>%
  summarize(avg_co2perfood = mean(co2perfood)) %>%
  ggplot(aes(x = reorder(food_category, avg_co2perfood), y = avg_co2perfood, fill = food_category)) + 
  geom_col() +
  coord_flip()+ 
  theme(legend.position = "none")
#This makes sense - animal based foods emit more co2
#it is interesting that lamb & goat is less "efficient" than beef
```

3. Scatterplot showing the average CO2 consumption by food category. A different way to view what we just saw in the bar plot. The ab-line represents complete efficiency, where the ratio between consumption and co2_emission is 1.

```{r}
#from video
food_consumption %>%
  group_by(food_category) %>%
  summarize(consumption = mean(consumption),
            co2_emission = mean(co2_emission)) %>%
  ggplot(aes(x=consumption, y = co2_emission, color = food_category))+
  geom_point()+
  geom_label_repel(aes(label=food_category, size=NULL), nudge_y=0.5)+
  geom_abline(slope =1)+ #1 assumes complete efficiency - calories in equals calories out
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
```


```{r, include=FALSE}
#from video
food_consumption %>%
  group_by(food_category) %>%
  summarize(consumption = mean(consumption),
            co2_emission = mean(co2_emission)) %>%
  mutate(residuals=co2_emission-consumption ) %>%
  ggplot(aes(x=consumption, y = co2_emission, color = food_category))+
  geom_point()+
  geom_label_repel(aes(label=food_category, size=NULL))+
  geom_segment(aes(xend=consumption, yend=consumption, x = consumption, y = co2_emission))+
  geom_abline(slope =1)+ #1 assumes complete efficiency - calories in equals calories out
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
#residuals show how energy inefficient some foods are (Lamb & Goat, Beef)
```

```{r}
#compare animal to non-animal products in terms of co2 emission

#food_consumption %>%
 # select(food_category) %>%
#  unique()
food_consumption <- food_consumption %>%
  mutate(food_type = if_else(food_category %in% c("Pork", "Poultry", "Beef", "Lamb & Goat", "Fish", "Eggs", "Milk - inc. cheese"), "Meat", "Vegan"))

food_consumption %>%
  group_by(food_type) %>%
  summarize(mean_consumption = mean(consumption),
                                    mean_co2 = mean(co2_emission))

#We can conduct a two sample t-test with equal variance to determine if there is a significant difference between the mean of consumption (or co2_emission) between food consumed that it meat-based and food consumed that is plant-based (vegan)
t.test(consumption ~ food_type, data=food_consumption, var.equal=TRUE)
#the result is a large p-value and similar means: mean consumption in meat group, 29.04; mean consumption in vegan group, 26.48.
t.test(co2_emission ~ food_type, data=food_consumption, var.equal=TRUE)
#The result is a very small p-value (less than 0.00) and very different means. mean co2-emission in meat group, 108.47, mean co2-emission in vegan group, 14.74. 
#These t-tests suggest that there is not a significant difference in consumption levels between meat and vegan groups but a very significant difference between the meat group and the vegan group in terms of co2 emissions.

food_consumption %>% 
  group_by(food_type) %>% 
  summarize(total_consumption = sum(consumption)) %>% table()
#With that said, in general, more vegan foods are consumed world-wide than meat foods by about 2:1 (13,769.92 meat, 26427.96 vegan)
```

```{r}
#what percentage of a country's food consumption is plant-based? 
#I used pivot_wider & created new variables 'meat_consumption' and 'plant_consumption' that group together consumption in these two categories, and then used these new totals to calculate this ratio
food_wider <- food_consumption %>%
  select(-co2_emission, -food_type) %>%
  pivot_wider(names_from = food_category, values_from = consumption) %>%
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE)))  %>%
  group_by(country) %>%
  mutate(meat_consumption = sum(pork + poultry + 
                                  beef + `lamb&goat` + 
                                  fish + eggs + `milk-inc.cheese`),
         plant_consumption = sum(wheatandwheatproducts + rice+
                                   soybeans+nutsinc.peanutbutter),
         lacto_ova_plant = sum(plant_consumption + eggs + `milk-inc.cheese`),
         ova_plant = sum(plant_consumption + eggs),
         perc_plantbased = round(plant_consumption/(meat_consumption+plant_consumption),2),
         perc_meatbased = round(meat_consumption/(meat_consumption+plant_consumption),2),
         perc_lacto_ova_plant = round(lacto_ova_plant/(meat_consumption+plant_consumption),2),
         perc_ova_plant = round(ova_plant/(meat_consumption+plant_consumption),2))

```

```{r}
#reactable of countries & perc plant or meat based
food_wider_table <- food_wider %>%
  select(country, perc_meatbased, perc_plantbased, perc_lacto_ova_plant)%>%
  arrange(desc(perc_meatbased))

reactable(food_wider_table)
```


```{r}
#top 5 countries for being plantbased
top5_plantbased <- food_wider %>%
  select(country, perc_plantbased) %>%
  arrange(desc(perc_plantbased)) %>%
  print(n=5)

kable(top5_plantbased[1:5, ],
      booktabs = T,
      col.names = c("Country","Percent of consumption plant-based, by kg per person"), 
      caption = "Top 5 Countries with the highest consumption of plant-based foods") %>% 
  footnote(general = "ADD",
           general_title = "Data Source") %>% 
  kable_styling(full_width = T)

#top 5 countries for being meatbased

top5_meatbased <- food_wider %>%
  select(country, perc_meatbased) %>%
  arrange(desc(perc_meatbased)) %>%
  print(n=5)

#top 5 by consumption of plant + eggs + cheese/milk
top5_lacto_ova <- food_wider %>%
  select(country, perc_lacto_ova_plant) %>%
  arrange(desc(perc_lacto_ova_plant)) %>%
  print(n=5)

#top 5 by consumption of plant + eggs
top5_ova_plant <- food_wider %>%
  select(country, perc_ova_plant) %>%
  arrange(desc(perc_ova_plant)) %>%
  print(n=5)

```

```{r}
#add a scatterplot where x=country, y=co2_emission, and size = consumption
#scatter <- food_consumption %>% group_by(country) %>%
 # ggplot(aes(x=consumption, y=co2_emission, fill=country))+ 
 
  #geom_point()+
  #scale_y_log10()+
  #scale_x_log10()+
  #theme(
   # legend.position="none",
   # axis.text.x = element_blank())

#ggplotly(scatter)
```


Around the world people eat more plant-based products (2:1) than meat. Plant-based products emit less co2 emissions than most meat-based products, especially beef, lamb, and goat. For some countries, most of their consumption is plant-based (i.e., Liberia's consumption is 82% plant based) while others, mostly animal-based (i.e., Finland, 86%). When we look at foods in a lacto-ova vegetarian diet (eggs, milk & cheese + plants), we find that India's consumption, for example, is 96% plant, egg, and dairy-based. If we go a step further and create a group representing an ova vegetarian diet (eggs + plants), Liberia's consumption of these products comes in the highest at 84%. Since eggs are an efficient food as well in terms of co2 emissions, comparing an ova-vegetarian group to lacto-ova vegetarian and vegan may be meaningful.

There are many limitations to this analysis. For one, food consumption ("consumption") is measured as kilograms of food per person per year. There could be some significant bias here, in that vegetables may weigh less that animal products. For example, the nutritional value of a kilogram of milk or cheese is very different from that of a kilogram of spinach. Second, analysis is done in a global accurate and does not take into account cultural differences across countries that account for population's food choices. For example, pork consumption is nil or very low in countries with large Muslim populations. 

Future analysis: Rather than looking at consumption measured in kilograms per person, a more meaningful analysis might be to consider consumption as calories consumed per person per year by food product. It would also be interesting to see what the relationship is between a country's GDP and the proportion of their diet that is plant-based. The intention behind this data may have been to shine light on the fact that animal products are less energy efficient than plant-based products. However, this propensity towards a plant-based diet may have more to do with a country's income than climate awareness.