---
title: "Food Consumption and CO2 Emissions, Tidy Tuesday"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

Sources

Data from Tidy Tuesday (<https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-18/readme.md>), February 18, 2020. I started by watching and following along with Andrew Couch's EDA on YouTube (<https://www.youtube.com/watch?v=VKCPYet9qLM>).

Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
library(tidytuesdayR)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(broom)
```

Data

```{r}
tuesdata <- tidytuesdayR::tt_load('2020-02-18')
tuesdata <- tidytuesdayR::tt_load(2020, week = 8)


food_consumption <- tuesdata$food_consumption %>%
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE))) %>%
rename(co2_emission = co2_emmission)

```

| Variable      | Class     | Description                       |
|---------------|-----------|-----------------------------------|
| country       | character | Country Name                      |
| food_category | character | Food Category                     |
| consumption   | double    | Consumption (kg/person/year)      |
| co2_emission  | double    | Co2 Emission (kg CO2/person/year) |

: Data Dictionary

Variables of interest

Review

```{r}
#130 countries and 11 food categories
food_consumption %>% count(country) 

#basic summary of the data set
food_consumption_NA <- food_consumption |>
  summarize(across(everything(), 
                   \(x) mean(is.na(x))))

#No NAs
food_consumption_0 <- 
  nrow(food_consumption[food_consumption$consumption == "0",]) #31 cases

#which countries/categories have cases where consumption = 0?
countries_0consumption <- food_consumption %>%
  filter(consumption == "0") %>%
  group_by(country) %>%
  print(n=31)
#food categories with a "0" value: Soybeans (22); Pork (8), and Lamb & Goat (1).
#Pork makes a lot of sense - countries with large Muslim populations are included in this list
food_consumption %>% 
  filter(consumption == "0" & food_category == "Pork") %>%
  group_by(country) %>%
  arrange(country) %>% 
  print()
#but what about soybeans? While I can believe that some of these countries eat very few soy products, I have a hard time believing they eat none. I am sceptical of these 0's.
food_consumption %>% 
  filter(consumption == "0" & food_category == "Soybeans") %>%
  group_by(country) %>%
  arrange(country) %>% 
  print(n=Inf)
```

Visualize

```{r}
#Distributions of CO2 and consumption are not as proportional as it appeared at first
options(scipen = 999)
food_consumption %>%
#tidy up the data before creating visual
  gather(key = "feature", value = "value", -country, -food_category) %>%
  ggplot(aes(x=feature, y = value, color = feature)) + 
  geom_boxplot() +
  geom_jitter(alpha = 0.1)+
  facet_wrap(~feature, scales = "fixed")+
  scale_y_log10()
```

```{r}
#Consumption = kg eaten by each person/year
#Co2 emission = kg of co2 produced by each person/year

#AC video question: for every kg of CO2 emitted, how many kgs of food do they eat? He's trying to determine which countries are more efficient. 
#of kg of food eaten produces how many kgs of CO2?

#Are more developed countries consuming more food / less efficient?

food_consumption %>%
  filter(consumption > 0 & co2_emission >0) %>%
  mutate(co2perfood = co2_emission/consumption) %>%
  group_by(food_category) %>%
  summarize(avg_co2perfood = mean(co2perfood)) %>%
  ggplot(aes(x = reorder(food_category, avg_co2perfood), y = avg_co2perfood, fill = food_category)) + 
  geom_col() +
  coord_flip()+ 
  theme(legend.position = "none")
#This makes sense - animal based foods emit more co2
#it is interesting that lamb & goat is less "efficient" than beef
```

```{r}
#scatterplot, consumption vs. co2_emission, grouped by food_category

food_consumption %>%
  filter(consumption > 0 & co2_emission >0) %>%
  mutate(co2perfood = co2_emission/consumption) %>%
  group_by(food_category) %>%
  summarize(avg_co2perfood = mean(co2perfood)) %>%
  ggplot(aes(x = reorder(food_category, avg_co2perfood), y = avg_co2perfood, fill = food_category)) + 
  scale_y_log10()+
  geom_point()+
  geom_label_repel(aes(label=food_category, size=NULL), nudge_y=0.5)+
  theme(
    legend.position= "none",
    axis.text.x = element_blank())+
  labs(x = "Food Categories", y = "Average CO2 Consumption per Food Category",
                  title = "Average Efficiency of Food Consumed by Food Category",
       subtitle = "data source: ")
  
```

```{r}
#What if we looked at emissions by type of diet? vegan vs. all

#first create indicator

```

```{r}
#from video
food_consumption %>%
  select(-country) %>%
  gather(key = "feature", value = "value", -food_category) %>%
  ggplot(aes(x = food_category, y = value, color = food_category)) +
  geom_boxplot() +
  coord_flip()+
  scale_y_log10()+
  facet_wrap(~feature, scales = "fixed")+
  theme(legend.position = "none")
```

```{r}
#from video
food_consumption %>%
  group_by(food_category) %>%
  summarize(consumption = mean(consumption),
            co2_emission = mean(co2_emission)) %>%
  ggplot(aes(x=consumption, y = co2_emission, color = food_category))+
  geom_point()+
  geom_label_repel(aes(label=food_category, size=NULL), nudge_y=0.5)+
  geom_abline(slope =1)+ #1 assumes complete efficiency - calories in equals calories out
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
```

```{r}
#from video
food_consumption %>%
  group_by(food_category) %>%
  summarize(consumption = mean(consumption),
            co2_emission = mean(co2_emission)) %>%
  mutate(residuals=co2_emission-consumption ) %>%
  ggplot(aes(x=consumption, y = co2_emission, color = food_category))+
  geom_point()+
  geom_label_repel(aes(label=food_category, size=NULL))+
  geom_segment(aes(xend=consumption, yend=consumption, x = consumption, y = co2_emission))+
  geom_abline(slope =1)+ #1 assumes complete efficiency - calories in equals calories out
  scale_x_log10()+
  scale_y_log10()+
  theme(legend.position = "none")
#residuals show how energy inefficient some foods are (Lamb & Goat, Beef)
```

```{r}
#top 5 consumers of each food category
#from video

food_consumption %>%
  group_by(food_category) %>%
  top_n(consumption, n = 5) %>% 
  arrange(food_category, -consumption) %>%
  ungroup() %>%
  count(country, sort = T)%>%
  filter(n != 1)
```

```{r}
#compare animal to non-animal products in terms of co2 emission
#from video
#food_consumption %>%
 # select(food_category) %>%
#  unique()

#food_consumption %>%
 # mutate(vegan = if_else(food_category %in% c("Wheat and Wheat Products", "Rice", "Soybeans", #"Nuts inc. Peanut Butter"), "Non Animal Product", "Animal Product"))%>%
  #count(food_category, vegan) %>%
 # select(consumption, co2_emission, vegan) %>%
#  gather(key = "type", value = "value", -vegan) %>%
 # mutate(type = as.factor(type), 
  #       vegan = as.factor(vegan)) %>%
  #group_by(type) %>%
  #do(test = t.test(value~vegan, data = (.))) %>%
  #tidy(test)%>%
  #print()
#didn't work for me, getting error
```

```{r}
#what percentage of a country's food consumption is plant-based? 
#I used pivot_wider & created new variables 'meat_consumption' and 'plant_consumption' that group together consumption in these two categories, and then used these new totals to calculate this ratio
food_wider <- food_consumption %>%
  select(-co2_emission) %>%
  pivot_wider(names_from = food_category, values_from = consumption) %>%
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE)))  %>%
  group_by(country) %>%
  mutate(meat_consumption = sum(pork + poultry + 
                                  beef + `lamb&goat` + 
                                  fish + eggs + `milk-inc.cheese`),
         plant_consumption = sum(wheatandwheatproducts + rice+
                                   soybeans+nutsinc.peanutbutter),
         lacto_ova_plant = sum(plant_consumption + eggs + `milk-inc.cheese`),
         perc_plantbased = round(plant_consumption/(meat_consumption+plant_consumption),2),
         perc_meatbased = round(meat_consumption/(meat_consumption+plant_consumption),2),
         perc_ova_plant = round(lacto_ova_plant/(meat_consumption+plant_consumption),2))
#top ten countries for being plantbased
food_wider %>%
  select(country, perc_meatbased, perc_plantbased) %>%
  arrange(desc(perc_plantbased)) %>%
  print(n=10)

#top ten countries for being meatbased

food_wider %>%
  select(country, perc_meatbased, perc_plantbased, perc_ova_plant) %>%
  arrange(desc(perc_meatbased)) %>%
  print(n=10)

#top ten by consumption of plant + eggs + cheese/milk
food_wider %>%
  select(country, perc_meatbased, perc_plantbased, perc_ova_plant) %>%
  arrange(desc(perc_ova_plant)) %>%
  print(n=10)
```

